---
title: "Group Analysis"
author: "Alex Schulte, Christine Lo, Matthew Reyes, Alexander Adia"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install.packages("corrplot")
library(tidyverse)
library(here)
library(ggplot2)
library(flextable)
library(arsenal)
library(corrplot)

```

## Research Question

Existing studies find that infants born at low birth weight (LBW) are at an increased risk of physical disabilities and impaired cognitive development. While genetic factors contribute to LBW, maternal smoking during pregnancy has been identified as the most significant modifiable risk factor. We seek to answer the following question: what is the effect of maternal smoking during pregnancy on the likelihood of having a LBW infant?

The target population for this study is live singleton first births in the US in 2015. We are limiting the population to singleton first births because multiples are associated with lower birth weight, and infants from subsequent pregnancies have been shown to have higher birth weights than those from first pregnancies.


## Target Causal Parameter

We aim to estimate the causal risk difference:
$\Psi^* (P^*) = P^* (Y1 - 1) - P^* (Y0 - 1)$ 
$= E^*(Y1) - E^*(Y0)$

The target causal parameter is the difference in the counterfactual risk of LBW if all expectant mothers in the population smoked during pregnancy vs. if all expectant mothers in the population did not smoke during pregnancy.

## Data Set

First, we import the data set for 2015 and inspect it, including variables available.

```{r import-data, echo = FALSE, message = FALSE, warning = FALSE}
# data set guide: https://data.nber.org/lbid/docs/LinkCO89Guide.pdf
# raw data files:
## download, unzip, and place in data folder:
## http://www.nber.org/lbid/1989/linkco1989us_den.csv.zip
## http://www.nber.org/lbid/1990/linkco1990us_den.csv.zip
## http://www.nber.org/lbid/1991/linkco1991us_den.csv.zip

here()
df <- read_csv(here("data/linkco2015usden.csv"), col_names = TRUE)
vars <- names(df)

vars
```

Next, we select variables of interest for our analysis by subsetting from the larger data set.

### Variables included in the dataset are as follows:
```{r select-vars, echo = FALSE, message = FALSE, warning = FALSE}

# subset the data to variables of interest
df_tobacco <- df %>%
  select(
    # W1 variables
    mrace15, mhisp_r, # race/ethnicity info
    mager9, # mother's age group
    dmar, # marital status
    meduc, # highest education level obtained of mother
    wic, # mother recipient of WIC
    mhtr, # mother height
    bmi, # mother bmi
    rf_pdiab, # pre-pregnancy diabetes (mother)
    rf_phype, # pre-pregnancy hypertension
    recare, # month prenatal care began
    # W2 variables
    wtgain, # mother weight gain during gestation
    rf_gdiab, # gestational diabetes
    ip_gon, ip_syph, ip_chlam, ip_hepb, ip_hepc, # infectious diseases present at birth
    # W3 variables
    oegest_comb, # weeks of gestation (17-47)
    dwgt_r, # mothers weight at time of birth
    sex, # sex of infant assigned at birth
    # Exposure variables
    cig0_r, cig1_r, cig2_r, cig3_r, # smoking status during each trimester
    # Outcome variables
    bwtr4, # outcome variable (infant birth weight)
    # subsetting/filtering variables
    tbo_rec, # birth order of infant
    illb_r, ilop_r, ilp_r # time since last live birth, last other pregnancy, last pregnancy
  )

```

Variable Name
Type
Descriptive summary of measure

smoked
Exposure (A, binary)
This variable is considered the intervention or exposure of interest - it’s a measure of whether the mother was considered a smoker (at least 1 cigarette/day) during any of the three trimesters.

lbw
Outcome (Y, binary)
This variable is the outcome, which is the weight of the infant at time of birth, classified as low birth weight (1) when the birthweight was below 2500 grams. Birth weight greater than 2500 grams is coded as 0.

mrace15
Endogenous covariate
Categorical race variable

mhisp_r
Endogenous covariate
Categorical variable indicating hispanic origin status

mager9
Endogenous covariate
Bins of age ranges

dmar
Endogenous covariate
Categorical variable of marital status

meduc
Endogenous covariate
Categorical variable of mother’s achieved education level

wic
Endogenous covariate
Indicator variable of mother receiving WIC benefits

mhtr
Endogenous covariate
Continuous variable of mother’s height

bmi
Endogenous covariate
Continuous variable of mother’s bmi

dwgt_r
Endogenous covariate
Continuous variable of mother’s weight at time of birth

rf_pdiab
Endogenous covariate
Indicator variable of mother having pre-pregnancy diabetes

rf_gdiab
Endogenous covariate
Indicator variable of mother with gestational diabetes

rf_phype
Endogenous covariate
Indicator variable of mother with pre-pregnancy hypertension

ip_gon
Endogenous covariate
Indicator variable of gonorrhea infection at time of birth

ip_syph
Endogenous covariate
Indicator variable of syphilis infection at time of birth

ip_chlam
Endogenous covariate
Indicator variable of chlamydia infection at time of birth

ip_hepb
Endogenous covariate
Indicator variable of Hepatitis B infection at time of birth

ip_hepc
Endogenous covariate
Indicator variable of Hepatitis C infection at time of birth

oegest_comb
Endogenous covariate
An edited obstetric estimate of weeks of gestation, discrete 17-47

tbo_rec
Endogenous covariate
Continuous variable of birth order

wtgain
Endogenous covariate
Continuous variable of mother’s weight gain during gestation

precare5
Endogenous covariate
Categorical variable of when prenatal care began

sex
Endogenous covariate
Categorical variable of sex of infant

### Data Cleaning

Then, we recode some of the variables of interest into outcome and exposure variables A and Y. We also prepare the covariates and endogenous variables for analysis by recoding them into indicator or dummy variables. We also remove missings or unknowns, which is a very conservative analysis approach - future analysis may utilize data imputation, but given the large number of records in this data set and the relatively small number of missing/unknown data, for the purpose of this assignment the more conservative approach is taken.

```{r recode-variables, echo = FALSE, message = FALSE, warning = FALSE}
# Removing unknowns/missings
df_noMissing <- df_tobacco %>%
  filter(
    bwtr4 != 4, # remove unknown birth weights
    cig0_r != 6, cig1_r != 6, cig2_r != 6, cig3_r != 6, # remove unknown smoking status at each trimester
    dwgt_r != 999, # remove unknown/unstated delivery weight of mother
    oegest_comb != 99,
    wic != "U",
    meduc != 9,
    mhtr != 99,
    dmar != 9,
    mhisp_r != 9,
    rf_pdiab != "U", rf_gdiab != "U", rf_phype != "U", ip_gon !="U", ip_syph != "U", ip_chlam != "U", ip_hepb != "U", ip_hepc != "U",
    ilop_r != 999, illb_r != 999, ilp_r != 999,
    mrace15 != 99,
    wtgain != 99,
    bmi != 99,
    tbo_rec != 9,
    recare != 99
  )

# #W1 variables
#     mrace15, mhisp_r, # race/ethnicity info
#     mager9, # mother's age group
#     dmar, # marital status
#     meduc, # highest education level obtained of mother
#     wic, # mother recipient of WIC
#     mhtr, # mother height
#     bmi, # mother bmi
#     rf_pdiab, # pre-pregnancy diabetes (mother)
#     rf_phype, # pre-pregnancy hypertension
#     recare, # month prenatal care began
# # W2 variables
#     wtgain, # mother weight gain during gestation
#     rf_gdiab, # gestational diabetes
#     ip_gon, ip_syph, ip_chlam, ip_hepb, ip_hepc, # infectious diseases present at birth
# # W3 variables
#     oegest_comb, # weeks of gestation (17-47)
#     dwgt_r, # mothers weight at time of birth
#     sex, # sex of infant assigned at birth
# # Exposure variables
#     cig0_r, cig1_r, cig2_r, cig3_r, # smoking status during each trimester
# # Outcome variables
#     bwtr4, # outcome variable (infant birth weight)
# # subsetting/filtering variables
#     tbo_rec, # birth order of infant
#     illb_r, ilop_r, ilp_r # time since last live birth, last other pregnancy, last pregnancy

firstBorns <- df_noMissing %>%
  filter(illb_r == 888)
  
# creating outcome and exposure variable
df_prep <- firstBorns %>%
  # add a Y outcome variable (lbw)
  mutate(lbw = ifelse(bwtr4 == 1 | bwtr4 == 2, 1, 0)) %>% 
  # create intervention variable A (smoked)
  mutate(smoked = ifelse(cig1_r >= 1 | cig2_r >= 1 | cig3_r >= 1, 1, 0)) %>%
  select(-bwtr4, -cig0_r, -cig1_r, -cig2_r, -cig3_r) # remove old variables used to create Y, A


# change variables of interest to indicator variables/factors for analysis
# helper function to create numerical indicator variables from Y/N coded variables
create_indicator <- function(df, variables){
  for (i in 1:length(variables)){
    df[[variables[i]]] <- ifelse(df[[variables[i]]] == "Y", 1, 0)
  }
  return(df)
}

convert <- c("wic", "rf_pdiab", "rf_gdiab", "rf_phype", "ip_gon", "ip_syph", "ip_chlam", "ip_hepb", "ip_hepc")
df_firstBorns <- create_indicator(df_prep, convert)

smoke_marg <- df_firstBorns %>%
  count(smoked)

lbw_marg <- df_firstBorns %>%
  count(lbw)

# number of records in dataset
n <- nrow(df_firstBorns)

# number of columns (variables) in dataset
n_cols <- length(df_firstBorns)
```

### Summary statistics of variables for analysis in the dataset

To better understand the data we're working with and get a sense for the distributions across variables (W1,W2, W3, A, Y), we can observe the information presented in Table 1 below.

```{r table1}
head(df_firstBorns)

summary_prep <- df_firstBorns %>%
  select(-illb_r, -ilop_r, -ilp_r)
summary_prep[,1:6] <- sapply(summary_prep[,1:6], as.factor)
summary_prep[,9:11] <- sapply(summary_prep[,9:11], as.factor)
summary_prep[,13:18] <- sapply(summary_prep[,13:18], as.factor)
summary_prep[,21:24] <- sapply(summary_prep[,21:24], as.factor)

summary_stats <- summarizor(summary_prep, by = "smoked", overall_label = "Overall")
summary_stats <- as_flextable(summary_stats, spread_first_col = TRUE, separate_with = "variable")
summary_stats <- labelizor(
  x = summary_stats,
  part = "header",
  labels = c(
    "0" = "No Smoking (A=0)", 
    "1" = "Smoked (A=1)"
  )
)

summary_stats <- labelizor(
  x = summary_stats,
  part = "body",
  j = c("sex"),
  labels = c(
    "sex" = "Sex of Child",
    "M" = "Male",
    "F" = "Female"
  )
)

summary_stats <- labelizor(x = summary_stats, part = "header", labels = stringr::str_to_title)

#### Uncomment to save table1 in various file formats ####

# save_as_docx("table1" = summary_stats, path = here("tables/table1.docx"))
# save_as_image(summary_stats, path = here("tables/table1.png"))
summary_stats
```

```{r collinearity}

corr_matrix <- df_firstBorns %>%
  select(-illb_r, -ilop_r, -ilp_r, -sex) %>%
  cor()

corrplot(corr_matrix, method = "number", order = "hclust", title = "Collinearity Matrix of Variables", bg ="white")

# lbw ~ oegest_comb,    -0.60
# dwgt_r ~ bmi,          0.84
# mager9 ~ wic,         -0.43
# mager9 ~ meduc,        

# we should keep bmi over dwgt_r, dwgt_r has less of an association to outcome (lbw) than bmi.

# arbitrary high correlation considered to be greater than 0.7 to 1 (or -0.7 to -1)
correlation <- as.data.frame(corr_matrix)
high_correlation <- correlation %>%
  filter_all(any_vars((0.5 < . & . < 1) | (-1 < . & . < -0.5)))
```

We should keep bmi over dwgt_r, dwgt_r has less of an association to outcome (lbw) than bmi.

```{r main-terms-model}

# Adjust for W1
model1 <- glm(lbw ~ mrace15+mhisp_r+mager9+dmar+meduc+wic+mhtr+bmi+rf_pdiab+rf_phype+recare, family = "binomial", data = df_firstBorns)
summary(model1)

```

The dataset consists of a number of variables describing births and pregnancies in the United States in the year 2015 with `r n` records after removing missing values, obtained from the National Center for Health Statistics.

### Marginal Distributions of Exposure and Outcome

For the mothers smoking status during the pregnancy, we observe:
`r smoke_marg`

For the low birth weight status at time of birth (outcome), we observe:
`r lbw_marg`

## Expected Challenges

Anticipated challenges include:

Identifying singleton births - we may need to create a unique identifier for each mother 
Computational strain given the size of the data 
Remaining potential for uncontrolled confounding (e.g., genetics, traumatic experiences during pregnancy)

## Expected Deviations

Potentially measuring only singleton births (single live birth per delivery) or even first live births from that mother.

## Analysis

### Sampling from Target Population

```{r sample-5percent}

set.seed(252)
n <- nrow(df_firstBorns)
sample_size <- n*0.05 # sample 5% of the larger dataframe, n = number of records in df

# take sample
df_5percent <- df_firstBorns[sample(n, size = sample_size, replace = FALSE),]

# create infectious disease indicator variable
df_5percent <- df_5percent %>% 
  # indicator variable for infectious disease present
  mutate(sti = ifelse(ip_gon == 1 | ip_syph == 1 | ip_chlam == 1 | ip_hepb == 1 | ip_hepc == 1, 1, 0)) %>%
  select(-ip_gon, -ip_syph, -ip_chlam, -ip_hepb, -ip_hepc, -illb_r, -ilop_r, -ilp_r)


write.csv(df_5percent, file=here("sample_5percent.csv"), row.names=FALSE)

```

### G-Comp
```{r g-comp}



```


### IPW
```{r IPW}


```


### SuperLearner/TMLE
```{r SL-setup}
set.seed(252)

# remove mediators from dataset
df <- df_5percent %>%
  select(-wtgain, -rf_gdiab, -sti, -oegest_comb, -dwgt_r, -sex)

# 63530 obs, so n/V = 63530/5 = 12706 obs per fold
n <- nrow(df)
V <- 5
n.per.fold <- n/V

fold <- NULL
for (k in 1:V) {
  fold <- c(fold, rep(k, n.per.fold))
}

# check to ensure we have 12706 per fold
# table(fold)
# head(fold)

# sampling
fold <- sample(fold)

# prediction and cross-validated matrices
pred <- matrix(NA, nrow = n, ncol = 4)
cv.risk <- matrix(NA, nrow = V, ncol = 4)

```

```{r SL-training}

for( v in 1:V ) {
  validation <- df[fold==v,]
  training <- df[fold!=v,]
  # QA check n/V = 12706
  # (nrow(validation))
  
  # fit estimators
  estA <- glm(lbw ~ ., family = "gaussian", data = training)
  estB <- glm(Y ~ mrace15, mhisp_r, )
  estB <- glm(Y ~ mrace15, mhisp_r, )
  estB <- glm(Y ~ mrace15, mhisp_r, )
}

```


## Works Cited

Almond, Douglas, Kenneth Y. Chay and David S. Lee. "The Costs Of Low Birth Weight," Quarterly Journal of Economics, 2005, v120(3,Aug), 1031-1083.

Bacci S, Bartolucci F, Chiavarini M, Minelli L, Pieroni L. Differences in birthweight outcomes: a longitudinal study based on siblings. Int J Environ Res Public Health. 2014 Jun;11(6):6472-84. doi: 10.3390/ijerph110606472. PMID: 25003169; PMCID: PMC4076673.

Bohn C, Vogel M, Poulain T et al.  Birth weight increases with birth order despite decreasing maternal pregnancy weight gain. Acta Paediatr 2021;110:1218–24.

National Center for Health Statistics (2015). Data File Documentations, Birth Cohort Linked Birth/Infant Death, 2015, National Center for Health Statistics, Hyattsville, Maryland. https://www.nber.org/research/data/linked-birthinfant-death-cohort-data

## Marginal Distribution Tables
```{r marginals, echo = FALSE, message = FALSE, warning = FALSE}
# helper function, create counts
counter <- function(df){
  for (k in 1:length(df)){
  result <- df %>%
      group_by(df[k]) %>%
      summarise(
        n = n()
      )
  print(result)
  }
}

# create tables of marginal distributions
# counter(df_firstBorns)
# 
# summary(df_firstBorns)

ggplot(data = df_firstBorns) +
  geom_histogram(aes(x=smoked))
ggplot(data = df_firstBorns) +
  geom_histogram(aes(x=lbw))
```
